Full-day, in home validation of infant body position measurements from inertial sensors

### 2023-02-28 manuscript.Rmd
intro

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home. Lightweight sensors (XX-XXg) can be embedded in garments to make recordings fully *mobile*, and they are *unobtrusive* because they do not require a researcher's presence nor do they record sensitive audio/video that might influence participants' behavior. Many commerically-available and inexpensive IMUs have > 12 hour battery life an onboard storage to record *real-time*, *full-day* data of movements at a sampling rate that is higher than typical video (e.g., 50-100 Hz). Moreover, past work has successfully recorded the rate of leg kicks [SMITH CITE] and activity intensity [CITES] in infants and children across an entire day.

The open question is whether *automatic classification* is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurement validity is acceptable over long recording periods. IMUs typically contain accelerometers that measure linear acceleration paired with gyroscopes that measure angular acceleration. Unlike motion tracking systems that might be used in a lab, IMUs do not provide data about the position of the body in space. Thus, data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration timeseries) into meaningful categories (e.g., supine, prone, sitting, upright). The difficulty of the classification task depends on the categories of interest. More basic aspects of movement, such as leg kicks or overall activity intensity, can be identified by taking the magnitude of acceleration (irrespective of direction) and applying thresholds or cut-points to define when a movement has occurred at a particular intensity.

Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is too complex for cut-point definitions to be accurate. First, the magnitude of movement can vary greatly *within* a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant and be stationary in "tummy time" or crawling in myriad ways [ADOLPH]. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a "V" position with legs fully extended, in a "W" position with knees bent. Sitting on a caregivers' lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates still more possibilities for how the arms and legs may move relative to the torso. Indeed, creating an all-encompassing set of rules for how to annotate sitting from video is no trivial task because of the various ways that sitting can occur in daily life.

Thus, modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults [CITEs], children [CITEs], and infants [CITEs]. Synchronized video with ground-truth human annotations creates training data for a machine learning algorithm, such as a random forest models, that can be later used to predict categories from future IMU data that was not annotated. Although in the current investigation we focus on classification of body position, this approach can be extended to categorize other aspects of movement, such as locomotion [CITE] and OTHER EXAMPLES? Our openly-shared dataset [CITE] contains synchronized video and IMU data, allowing researchers to annotate and train models for behaviors beyond body position.

REVIEW PAST WORK IN BODY POSITION CLASSIFICATION, END WITH WHAT IS MISSING.

>>>>

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home. Lightweight sensors (XX-XXg) can be embedded in garments to make recordings fully *mobile*, and they are *unobtrusive* because they do not require a researcher's presence nor do they record sensitive audio/video that might influence participants' behavior [(Cliff et al., 2009; de Barbaro, 2019; Lobo et al., 2019; Bruijns et al., 2020)]. Many commerically-available and inexpensive IMUs have > 12 hour battery life an onboard storage to record *real-time*, *full-day* data of movements at a sampling rate that is higher than typical video (e.g., 50-100 Hz). Moreover, past work has successfully recorded the rate of leg kicks [(Smith et al., 2017; Jiang et al., 2018)] and activity intensity [Hagenbuchner et al., 2015; Trost et al., 2018] in infants and children across an entire day. 
<!-- CHECK THAT CITATIONS ARE FULL DAY -->

The open question is whether *automatic classification* is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurement validity is acceptable over long recording periods. IMUs typically contain accelerometers that measure linear acceleration paired with gyroscopes that measure angular acceleration. Unlike motion tracking systems that might be used in a lab, IMUs do not provide data about the position of the body in space. Thus, data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration timeseries) into meaningful categories (e.g., supine, prone, sitting, upright). The difficulty of the classification task depends on the categories of interest. More basic aspects of movement, such as overall activity intensity, can be identified by taking the magnitude of acceleration (irrespective of direction) and applying thresholds or cut-points to define when a movement has occurred at a particular intensity [(Trost et al., 2012; Kuzik et al., 2015; Hager et al., 2017; Armstrong et al., 2019)].

Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is too complex for cut-point definitions to be accurate. First, the magnitude of movement can vary greatly *within* a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant and be stationary in "tummy time" or crawling in myriad ways [ADOLPH]. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a "V" position with legs fully extended, in a "W" position with knees bent. Sitting on a caregivers' lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates still more possibilities for how the arms and legs may move relative to the torso. Indeed, creating an all-encompassing set of rules for how to annotate sitting from video is no trivial task because of the various ways that sitting can occur in daily life. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity [Kwon]. 
GREENSPAN...

Thus, modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults [(Preece et al., 2009; Arif and Kattan, 2015)], children [(Nam and Park, 2013; Zhao et al., 2013; Ren et al., 2016; Stewart et al., 2018)], and infants [Franchak; Airaksinen; Yao]. Synchronized video with ground-truth human annotations creates training data for a machine learning algorithm, such as a random forest models, that can be later used to predict categories from future IMU data that was not annotated. Although in the current investigation we focus on classification of body position, this approach can be extended to categorize other aspects of movement, such as locomotion [CITE] and OTHER EXAMPLES? Accompanying our paper, we have created a database [CITE] containing synchronized video and IMU data, allowing researchers to annotate and train models for behaviors beyond body position.

Three prior investigations have used machine learning to categorize infant body position from IMUs towards the goal of collecting full-day data, and each has yielded encouraging validation results. @AiraksinenRasanen2020 tested 4- to 8-month-olds in a laboratory visit, and found 95% accuracy in distinguishing between body position categories. Using a wider age range of XX-XX, [FRANCHAK] found XX% accuracy in a laboratory validation study in categorizing body position.
